---
title: "Getting started with ShinyPopGen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ShinyPopGen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

**ShinyPopGen** is an interactive Shiny application for exploratory and
descriptive population genetics analyses from individual-based multilocus
datasets (microsatellites, codominant markers).

It covers the full analytical workflow:

1. Data import and automatic column detection
2. Allele frequency distributions and missing data
3. General statistics (N, Na, Ho, He, F-statistics)
4. Local panmixia -- FIS per locus and population (Weir & Cockerham 1984)
5. Global panmixia -- FIT across all populations
6. Population subdivision -- FST per locus and globally
7. Genetic diversities -- HS, HT, locus bootstrap
8. Linkage disequilibrium -- pairwise G-test with permutation p-values
9. Null allele frequency estimation

## Launch the application

```{r launch}
library(shinypopgen)
run_app()
```

The application opens in your default browser. All computations run locally;
no data is sent to any server.

## Data format

ShinyPopGen accepts tab-separated or comma-separated plain text files.
Two genotype encodings are supported.

### Single-column format

Each locus occupies one column; alleles are separated by a character
(`/`, `-`, `_`, etc.):

```
Population  B12      C07      D12
Pop1        192/194  145/192  92/112
Pop1        200/200  179/179  108/112
Pop2        145/145  92/92    92/100
```

### Two-column format

Each locus occupies two consecutive columns (one per allele):

```
Population  B12  B12  C07  C07
Pop1        192  194  145  192
Pop1        200  200  179  179
Pop2        145  145   92   92
```

Missing genotypes should be coded consistently (default: `0`).

## Worked example: *Boophilus* cattle ticks

The package ships a demonstration dataset from Thierry de Meeus' teaching
material (`inst/extdata/default_dataset.csv`): 698 adult *Boophilus* ticks
sampled from cattle across New Caledonia, genotyped at 6 microsatellite loci
(B12, C07, D12, D10, A12, C03).

### 1. Load the default dataset

Click **"Load Default Data"** in the sidebar of the *Data Import & Filtering*
tab. The application will:

- Import the file via DuckDB
- Auto-detect the population column (`Pop`) and GPS coordinates
- Detect the 12 allele columns (two-column format, loci B12 -- C03)
- Build the internal `hf` table (individual x locus packed integers)
- Display a preview and map of sampling localities

### 2. Allele frequencies

Navigate to *Allele Freq & Missing Data*. You will see per-locus allele
frequency bar charts and a missing-data summary table. The Boophilus dataset
has no missing data.

### 3. General statistics

The *General Statistics* tab shows sample size (N), number of alleles (Na),
observed heterozygosity (Ho), and expected heterozygosity (He) per locus and
population, together with overall Weir & Cockerham (1984) F-statistics.

### 4. Local panmixia (FIS)

The *Local Panmixia* tab reports per-locus FIS values within each population,
with:

- **Bootstrap 95% CI** (individual resampling, 1000 replicates by default)
- **Permutation p-value** (within-population allele shuffling,
  1000 permutations; `p = (B_ge + 1) / B`)

### 5. Linkage disequilibrium

The *Linkage Disequilibrium* tab tests all locus pairs within each population.
The G-statistic is:

```
G = 2 * sum(O * log(O / E))
```

where O and E are observed and expected counts under independence.
P-values are Monte Carlo estimates from Fisher-Yates permutations of
one locus column within each population.

### 6. Export results

Every results table has a download button (.csv / .txt). The formatted
dataset can also be downloaded from the *Data Import* tab.

## Statistical methods

All F-statistics follow **Weir & Cockerham (1984)**:

> Weir, B.S. & Cockerham, C.C. (1984). Estimating F-statistics for the
> analysis of population structure. *Evolution*, **38**, 1358--1370.

Permutation p-values use the formula `p = (n_ge + 1) / B` where `n_ge` is
the number of permuted statistics >= the observed value and `B` is the number
of permutations, following:

> Phipson, B. & Smyth, G.K. (2010). Permutation p-values should never be
> zero: calculating exact p-values when permutations are randomly drawn.
> *Statistical Applications in Genetics and Molecular Biology*, **9**, 39.

Heavy computations (bootstrap, permutation loops) are parallelised with
OpenMP via C++17 (Rcpp).

## Session info

```{r session}
sessionInfo()
```
