---
title: "Getting started with ShinyPopGen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ShinyPopGen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

**ShinyPopGen** is an interactive Shiny application for exploratory and
descriptive population genetics analyses from individual-based multilocus
datasets (microsatellites, codominant markers).

It covers the full analytical workflow:

1. Data import and automatic column detection
2. Allele frequency distributions and missing data
3. General statistics (N, Na, Ho, He)
4. Null allele frequency estimation (FreeNA EM algorithm, ENA correction)
5. Local panmixia -- FIS per locus and population (Weir & Cockerham 1984)
6. Global panmixia -- FIT across all populations
7. Population subdivision -- FST per locus and globally
8. Genetic diversities -- HS, HT, with resampling over individuals,
   populations, or loci
9. Linkage disequilibrium -- pairwise G-test with permutation p-values
10. Isolation by distance -- FST/(1-FST) vs. geographic distance, Mantel test

## Launch the application

```{r launch}
library(shinypopgen)
run_app()
```

The application opens in your default browser. All computations run locally;
no data is sent to any server.

## Data format

ShinyPopGen accepts CSV or tabulation-delimited TXT files. The separator
(comma, semicolon, or tab) is detected automatically. The file must contain
a header row, one *population* column, and one or more *genetic marker*
columns; metadata columns (e.g. individual ID, latitude, longitude) are
optional and preserved through the analysis. Two genotype encodings are
supported, but a single file must use only one of them.

### Single-column format

Each locus occupies one column; alleles are separated by a character
(`/`, `-`, `_`, etc.):

```
Population  B12      C07      D12
Pop1        192/194  145/192  92/112
Pop1        200/200  179/179  108/112
Pop2        145/145  92/92    92/100
```

### Two-column format

Each locus occupies two columns, one per allele. The first allele column
carries the locus name (e.g. `B12`); the second must be named `B12_1` (or
`B12.1`) -- any column whose name ends in a `_1`/`.1`-`.9` suffix is
auto-paired with the column obtained by removing that suffix. Paired
columns do not need to be adjacent:

```
Population  B12  B12_1  C07  C07_1
Pop1        192  194    145  192
Pop1        200  200    179  179
Pop2        145  145     92   92
```

If auto-detection picks up the wrong columns, use the manual column
assignment panel on the Import page instead.

Missing genotypes are coded as `0` (or `0/0`) by default and are
customisable on import. For **null allele** screening specifically, null
homozygotes must be coded as `999999/999999`, distinct from missing data
(`0/0`); a per-locus override panel is available in the Null Alleles tab.

## Worked example: *Boophilus* cattle ticks

The package ships a demonstration dataset from Thierry de Meeûs' teaching
material (`inst/extdata/default_dataset.csv`): 698 adult *Boophilus* ticks
sampled from cattle across New Caledonia, genotyped at 6 microsatellite loci
(B12, C07, D12, D10, A12, C03).

The steps below follow the app's recommended workflow (see the in-app
**Help** tab).

### 1. Import the data

Click **"Load Default Data"** in the sidebar of the *Import Data* tab. The
application will:

- Import the file via DuckDB
- Auto-detect the population column (`Pop`) and GPS coordinates
- Detect the 12 allele columns (two-column format, loci B12 -- C03)
- Build the internal genotype table (individual x locus, packed integers)
- Display a preview table and a map of sampling localities

### 2. Allele frequencies

Navigate to *Allele Freq*. You will see per-locus allele frequency bar
charts and a missing-data summary table; flag any locus with more than
about 20% missing data. The Boophilus dataset has no missing data.

### 3. General statistics

The *General Stats* tab reports sample size (N), number of alleles (Na),
observed heterozygosity (Ho), and expected heterozygosity (He) per locus
and population.

### 4. Null alleles

If the Ho/He ratio at a locus looks unusually low, check the *Null Alleles*
tab: null allele frequencies are estimated per locus x population by the
FreeNA EM algorithm (Chapuis & Estoup 2007), with both raw and
ENA-corrected FST reported.

### 5. Local panmixia (FIS)

The *Local Panmixia* tab reports per-locus FIS values within each
population (Weir & Cockerham 1984), with:

- **Bootstrap 95% CI** (resampling over individuals; 5,000 replicates by
  default)
- **Permutation p-value** (alleles reassigned within each population;
  5,000 permutations by default)

### 6. Global panmixia (FIT)

The *Global Panmixia* tab reports the multilocus FIT across all
populations, with the same bootstrap/permutation approach as above.

### 7. Subdivision (FST)

The *Subdivision* tab estimates FST (Weir & Cockerham 1984) per locus and
overall, with population-block or locus bootstrap CI and a permutation
p-value; a genotype-reassignment G-test (as in FSTAT, Goudet et al. 1996)
is also available for individuals with a complete multilocus genotype.

### 8. Genetic diversities (HS, HT)

The *Diversities* tab reports HS (within-population) and HT (total) gene
diversity (Nei 1987), per locus and as multilocus estimates, with
confidence intervals from resampling over individuals, populations, or
loci depending on the estimator.

### 9. Linkage disequilibrium

The *LD* tab tests all locus pairs within each population. The G-statistic
is:

```
G = 2 * sum(O * log(O / E))
```

where O and E are observed and expected counts under independence.
P-values are Monte Carlo estimates from permutations of one locus column
within each population (10,000 permutations by default).

### 10. Isolation by distance

The *Isolation by Distance* tab requires GPS metadata (present in the
Boophilus dataset). It plots pairwise FST/(1-FST) against geographic
distance following **Rousset (1997)**, and offers a Mantel permutation test
of the association.

### Export results

Every results table has a download button (.csv / .txt). The formatted
dataset can also be downloaded from the *Import Data* tab.

## Statistical methods

All F-statistics follow **Weir & Cockerham (1984)**:

> Weir, B.S. & Cockerham, C.C. (1984). Estimating F-statistics for the
> analysis of population structure. *Evolution*, **38**, 1358--1370.

Gene diversity (HS, HT) follows **Nei (1987)**:

> Nei, M. (1987). *Molecular Evolutionary Genetics*. Columbia University
> Press, New York.

Null allele frequencies and the ENA correction follow **Chapuis & Estoup
(2007)** (the FreeNA method):

> Chapuis, M.P. & Estoup, A. (2007). Microsatellite null alleles and
> estimation of population differentiation. *Molecular Biology and
> Evolution*, **24**, 621--631.

Isolation by distance follows **Rousset (1997)**:

> Rousset, F. (1997). Genetic differentiation and estimation of gene flow
> from F-statistics under isolation by distance. *Genetics*, **145**,
> 1219--1228.

The genotype-reassignment G-test for subdivision follows **FSTAT**:

> Goudet, J., Raymond, M., de Meeus, T. & Rousset, F. (1996). Testing
> differentiation in diploid populations. *Genetics*, **144**, 1933--1940.

Permutation p-values use the formula `p = (b + 1) / (m + 1)`, where `b` is
the number of permuted statistics at least as extreme as the observed value
and `m` is the number of permutations, following:

> Phipson, B. & Smyth, G.K. (2010). Permutation p-values should never be
> zero: calculating exact p-values when permutations are randomly drawn.
> *Statistical Applications in Genetics and Molecular Biology*, **9**, 39.

Heavy computations (bootstrap, permutation loops) are parallelised with
OpenMP via C++17 (Rcpp).

## Session info

```{r session}
sessionInfo()
```
